---
output:
  pdf_document: 
    keep_tex: yes
    fig_caption: yes
    number_sections: yes
    toc: yes
geometry: margin=2.54cm
title: "Seasonal Salt Content in the Neuse River"
subtitle: "https://github.com/atf35/CarpenterFischerMacDonald_ENV872_EDA_FinalProject"
author: "Atalie Fischer, Kathlyn MacDonald, Jack Carpenter"
fontsize: 12pt
mainfont: Times New Roman
  
---

\newpage



\listoftables 
Table 1: Summary of the data used..........................................6

Table 2: P-values for Seasonal Mann-Kendall tests...........................11
\newpage

\listoffigures 
\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(include = FALSE)
# Set your working directory
# setwd("/Users/kmac9/OneDrive/Documents/Duke/Year 1/Spring 2022/ENV Data/CarpenterFischerMacDonald_ENV872_EDA_FinalProject/")
# setwd("/Users/ataliefischer/Desktop/EDA/CarpenterFischerMacDonald_ENV872_EDA_FinalProject/")
getwd()

#install.packages("fable")
#install.packages("tseries")

# Load packages
library(tidyverse)
library(lubridate)
library(trend)
library(forecast)
library(dataRetrieval)
library(sf)
library(maps)
library(knitr)
library(nhdplusTools)
library(cowplot)
#library(fable)
library(tseries)
#install.packages("reshape2")
library(reshape2)

# Set ggplot theme
mytheme <- theme_classic(base_size = 10) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "top")

## Load datasets

#load water quality data for the Neuse River at Kinston, NC
NeuseWQ <- readWQPqw(siteNumbers = "USGS-02089500", # Neuse River at Kinston, NC
                     parameterCd = c("00915", # calcium, filtered mg/L
                                     "00925", # magnesium, filtered mg/L
                                     "00930", # sodium,filtered mg/L
                                     "90095"),  # Specific conductance, uS/cm
                     startDate = "1976-01-01",
                     endDate = "")
#write.csv(NeuseWQ, row.names = TRUE, file = "../Data/Raw/NeuseWQ_raw.csv")

#load discharge data for Neuse River at Kinston, NC
NeuseFlow <- readNWISdv(siteNumbers = "02089500", # Neuse River at Kinston, NC
                     parameterCd = "00060", # discharge (cfs)
                     startDate = "1976-01-01",
                     endDate = "")
#write.csv(NeuseFlow, row.names = TRUE, file = "../Data/Raw/NeuseFlow_raw.csv")
```

# Rationale and Research Questions
The maintenance of healthy and functioning water systems is critical not just to human life, but also to the survival of countless other species and their interconnected ecosystems. Conductivity, or the ability of water to pass an electrical current via dissolved salts and other minerals, is a strong indicator of salinity, a key component of water quality. Changes in conductivity over time suggest potential salt and mineral pollutants entering the system, and also potential salinization over time. 

<<<<<<< HEAD
For this study, we will analyze how specific conductance varies over time. In particular, we will address the impact of winterizing the roadways with salt on water quality. Our main research question is: *Is salting the roads a main driver in changes to water conductivity, or are other  minerals a significant factor?* Here, we chose to analyze water quality and flow data from the Neuse River in Kinston, North Carolina from 1976 to 2022. We chose this river due to familiarity with the data and its proximity to an urban center (Kinston pop = 20,398).
=======
=======
For this study, we will analyze how specific conductance varies over time. In particular, we will address the impact of winterizing the roadways with salt on water quality. Our main research question is: Is salting the roads a main driver in changes to water conductivity, or are other  minerals a significant factor? Here, we chose to analyze water quality and flow data from the Neuse River in Kinston, North Carolina from 1976 to 2022. We chose this river due to familiarity with the data and its proximity to an urban center (Kinston pop = 20,398). We also know from living in the headwaters of this river that salt is applied to roadways in the upper watershed in the winter (albeit infrequently). Since the gage at Kinston is further downstream,  we will examine whether there are significant downstream effects of road salting on water quality in the Neuse River basin. ADD MORE?

>>>>>>> 35fb7013fec92c5bba2e868721ba451ca364fff1

We used the following research questions to guide our work: 

> 1. How does specific conductance vary over time and seasonally? 
> 2. Is calcium, magnesium, or sodium the driver of specific conductance?


\newpage

# Dataset Information

**Neuse River water quality and discharge data at Kinston, North Carolina.** The gage information comes from the United States Geologic Survey (USGS) National Water Information Systems (NWIS) database. USGS gage stations typically collect discharge and a subset collect water quality data. This water quality data may include nutrient concentrations, concentrations of chlorophyll a, specific conductivity, and concentrations of certain ions. Since seasonal salinity trends and their potential sources are the focus of this study, the water quality data being examined includes specific conductivity and concentrations of calcium, magnesium, and sodium in the water column. Specific conductivity will be used as a proxy for salinization, and the relative amounts of each salt ion will be examined in the hopes of identifying a potential source of any seasonal salinity increases, in particular the contribution of road salts to salinization. 
 
To obtain both sets of data, the dataRetrieval package was used to connect directly to the NWIS database and pull water quality and discharge data without needing to download any files. Both sets of data are pulled starting in 1976 and end at the most recent data point in the database in 2022. 

Table 1: Summary of the data used.

Dataset           | Info
------------------|------------------------
NeuseWQ           | Water quality data collected at USGS gage 02089500
NeuseFlow         | Discharge data collected at USGS gage 02089500
Retrieved from    | USGS NWIS database with dataRetrieval package
Variables         | Specific conductivity, calcium, sodium, & magnesium concentrations, sample date, discharge
Date Range        | 1976 through 2022, wrangled to 2013 through 2022



\newpage

# Exploratory Analysis

The first step we took in our initial exploratory analysis was to wrangle the water quality (WQ) dataset to include only the columns of interest. This included the sampling dates and concentrations for specific conductance, calcium (Ca), sodium (Na), and magnesium (Mg), which were each given separate columns. This dataset contains monthly observations, however, not necessarily sampled on the first of each month. We wrangled the WQ dataset to round the dates to the first of the month to ensure that there are evenly spaced time steps across the years, a necessary condition for time series analyses. 

```{r WQ wrangling, include=FALSE}
# select columns of interest
NeuseWQ_processed <- NeuseWQ %>%
  select(ActivityStartDate, HydrologicCondition, CharacteristicName, ResultMeasureValue) %>%
  mutate(Variable = case_when(CharacteristicName == "Calcium" ~ "Ca_mg.L",
                              CharacteristicName == "Magnesium" ~ "Mg_mg.L",
                              CharacteristicName == "Sodium" ~ "Na_mg.L",
                              CharacteristicName == "Specific conductance" ~ "SpC_uS.cm")) %>%
  select(-CharacteristicName) %>%
  pivot_wider(names_from = "Variable", values_from = "ResultMeasureValue") 

# create sequence of dates with even time steps 
Months <- data.frame(Date_monthrounded = seq.Date(from = as.Date("1976-01-01"), to = as.Date("2022-12-01"), by = "month"))

# round sample dates down to the first of the month to ensure even time steps
NeuseWQ_processed <- NeuseWQ_processed %>% 
  mutate(Year = year(ActivityStartDate),
         Month = month(ActivityStartDate),
         Date_monthrounded = floor_date(ActivityStartDate, "month")) %>% # round date
  arrange(ActivityStartDate)

# join the WQ and months together to find gaps in data
NeuseWQ_monthly <- left_join(Months, NeuseWQ_processed)

# average months to consolidate months with multiple observations
NeuseWQ_monthly <- NeuseWQ_monthly %>% 
  group_by(Date_monthrounded) %>% 
  summarise(AvgCa_mg.L = mean(Ca_mg.L),
            AvgMg_mg.L = mean(Mg_mg.L),
            AvgNa_mg.L = mean(Na_mg.L),
            AvgSpC_uS.cm = mean(SpC_uS.cm))

#pivot longer
NeuseWQ_processed_pivotLonger <- pivot_longer(NeuseWQ_processed, Mg_mg.L:Na_mg.L, names_to = "Mineral", values_to = "ResultMeasureValue")


# write csv file for WQ data
#write.csv(NeuseWQ_processed, row.names = TRUE, file = "./Data/Processed/NeuseWQ_processed.csv")
#write.csv(NeuseWQ_monthly, row.names = TRUE, file = "./Data/Processed/NeuseWQ_monthly.csv")
```

We plotted the specific conductance over time to visualize any gaps in our dataset (Figure 1). We see that the WQ dataset contains many long periods of missing data for specific conductance. Since these missing periods frequently span across many years, we chose to look at WQ data from 2013 through 2021 (Figure 2). There are no missing data points from this period of time, and will therefore not require any interpolation of this dataset. 

```{r Visualize monthly WQ data, echo=FALSE, fig.align='left', fig.cap="Specific Conductance in the Neuse River", warning=FALSE, include=TRUE}
ggplot(NeuseWQ_monthly) +
  geom_line(aes(x = Date_monthrounded, y = AvgSpC_uS.cm)) +
  labs(x = "Date", y = expression("Specific Conductance ("*mu*"S/cm)"))
```

```{r Visualize 2013-2021 WQ data (Conductance), include=TRUE, echo=FALSE, warning=FALSE, fig.align='left', fig.cap="Specific Conductance in the Neuse River from 2013 through 2021."}
NeuseWQ_short <- NeuseWQ_monthly %>% 
  filter(Date_monthrounded >= "2013-01-01" & Date_monthrounded < "2022-01-01")

SpCplot <- ggplot(NeuseWQ_short) +
  geom_line(aes(x = Date_monthrounded, y = AvgSpC_uS.cm)) +
  labs(x = "Date", y = expression("Specific Conductance ("*mu*"S/cm)"))
SpCplot
```

```{r Water Quality Dataset Summary, include=FALSE, echo=FALSE}
summary(NeuseWQ_short)
```


\newpage

We are also interested in the flow of the Neuse River because this factor may affect salinity. For example, higher discharges may dilute any salinity and drier periods may reflect higher salt content. We started by wrangling the flow dataset to include the parameters of interest, sampling date and discharge  (Figure 3). 

```{r Flow wrangling, include=FALSE}
# flow dataset
NeuseFlow_processed <- NeuseFlow %>%
  select(ActivityStartDate = Date, discharge_cfs = X_00060_00003) %>% 
  filter(ActivityStartDate >= "2013-01-01" & ActivityStartDate <= "2022-01-01") %>% 
  mutate(Year = year(ActivityStartDate),
         Month = month(ActivityStartDate))

## check even time steps
Days <- data.frame(Days = seq.Date(from = as.Date("2013-01-01"), to = as.Date("2022-01-01"), by = "day"))
NeuseFlow_days <- left_join(Days, NeuseFlow_processed, by = c("Days" = "ActivityStartDate"))
summary(NeuseFlow_days) # no NA's in dataset

# write files to csv
#write.csv(NeuseFlow_processed, row.names = TRUE, file = "./Data/Processed/NeuseFlow_processed.csv")
#write.csv(NeuseFlow_days, row.names = TRUE, file = "./Data/Processed/NeuseFlow_days.csv")
```

```{r Discharge summary, include=FALSE, echo=FALSE, message=FALSE}
summary(NeuseFlow_processed) #no NAs in the dataset
```

```{r Discharge over time, include=TRUE, echo=FALSE, message=FALSE, fig.align='left', fig.cap="Discharge in the Neuse River from 2013 through 2021."}
discharge_plot <- ggplot(NeuseFlow_processed) +
  geom_line(aes(x = ActivityStartDate, y = discharge_cfs)) +
  labs(x = "Date", y = "Discharge (cfs)") +
  scale_y_log10()
discharge_plot
```

\newpage

# Analysis


## Question 1: How does specific conductance vary over time and seasonally?


### Is the specific conductivity of the Neuse River monotonically changing over time?

To determine whether specific conductance varies over time, we first constructed and decomposed the time series data. The time series decomposition of specific conductivity shows a seasonal component in the Neuse River (Figure 4). To test whether the Neuse River exhibits a monotonic trend in specific conductance, we conducted a seasonal MannKendall test. With a p-value > 0.05, we accept the null hypothesis that there the specific conductance of the Neuse River exhibits no monotonic trend (Table 2), meaning that specific conductivity is neither increasing or decreasing over time (Figure 5).

```{r Conductivity time series and decomp, include=FALSE}
#Creating Time Series for conductivity
NeuseCond_ts <- ts(NeuseWQ_short[[5]], frequency = 12)
NeuseCond_decomp <- stl(NeuseCond_ts, s.window = "periodic")
```

```{r Plot of Conductivity Time Series Decomp, include=TRUE, echo=FALSE, message=FALSE, fig.align='center', fig.width = 6, fig.height = 4, fig.cap="Time Series Decomposition of Conductivity in the Neuse River."}
plot(NeuseCond_decomp, main = "Specific Conductivity")
```

```{r Plot of Conductivity over Time with Trend, include=TRUE, echo=FALSE, message=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap="Specific Conductance in the Neuse River does not exhibit a monotonic trend over time."}
ggplot(NeuseWQ_short, aes(x = Date_monthrounded, y = AvgSpC_uS.cm)) +
  geom_point() +
  geom_line() +
  xlab("Date") +
  ylab(expression("Average Specific Conductance ("*mu*"S/cm)")) +
  geom_smooth(method = lm)
```

\newpage

Unsurprisingly, the time series decomposition and seasonal MannKendall tests of calcium (Ca) (Figures 6-7), magnesium (Mg) (Figures 8-9), and sodium (Na) (Figures 10-11) show that the these minerals also do not  exhibit monotonic trends over time (p-values in Table 2), meaning that mineral content is also neither increasing or decreasing over time. 

```{r Salt time series and decomps, include=FALSE}
#Decomposing by Mineral
#Calcium
NeuseCond_Calcium_ts <- ts(NeuseWQ_short$AvgCa_mg.L, frequency = 12)
NeuseCond_Calcium_decomp <- stl(NeuseCond_Calcium_ts, s.window = "periodic")

#Magnesium
NeuseCond_magnesium_ts <- ts(NeuseWQ_short$AvgMg_mg.L, frequency = 12)
NeuseCond_magneisum_decomp <- stl(NeuseCond_magnesium_ts, s.window = "periodic")

#Sodium
NeuseCond_sodium_ts <- ts(NeuseWQ_short$AvgNa_mg.L, frequency = 12)
NeuseCond_sodium_decomp <- stl(NeuseCond_sodium_ts, s.window = "periodic")
```

```{r Plot of Calcium Time Series Decomp, include=TRUE, echo=FALSE, message=FALSE, fig.align='center', fig.width = 6, fig.height = 4, fig.cap="Time Series Decomposition of Calcium in the Neuse River."}
plot(NeuseCond_Calcium_decomp, main = "Calcium")
```

```{r Plot of Calcium over Time with Trend, include=TRUE, echo=FALSE, message=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap="Calcium in the Neuse River does not exhibit a monotonic trend over time."}
ggplot(NeuseWQ_short, aes(x = Date_monthrounded, y = AvgCa_mg.L)) +
  geom_point() +
  geom_line() +
  xlab("Date") +
  ylab("Average Calcium (mg/L)") +
  geom_smooth(method = lm)
```

```{r Plot of Magnesium Time Series Decomp, include=TRUE, echo=FALSE, message=FALSE, fig.align='center', fig.width = 6, fig.height = 4, fig.cap="Time Series Decomposition of Magnesium in the Neuse River."}
plot(NeuseCond_magneisum_decomp, main = "Magnesium")
```

```{r Plot of Magnesium over Time with Trend, include=TRUE, echo=FALSE, message=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap="Magnesium in the Neuse River does not exhibit a monotonic trend over time."}
ggplot(NeuseWQ_short, aes(x = Date_monthrounded, y = AvgMg_mg.L)) +
  geom_point() +
  geom_line() +
  xlab("Date") +
  ylab("Average Magnesium (mg/L)") +
  geom_smooth(method = lm)
```

```{r Plot of Sodium Time Series Decomp, include=TRUE, echo=FALSE, message=FALSE, fig.align='center', fig.width = 6, fig.height = 4, fig.cap="Time Series Decomposition of Sodium in the Neuse River."}
plot(NeuseCond_sodium_decomp, main = "Sodium")
```

```{r Plot of Sodium over Time with Trend, include=TRUE, echo=FALSE, message=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap="Sodium in the Neuse River does not exhibit a monotonic trend over time."}
ggplot(NeuseWQ_short, aes(x = Date_monthrounded, y = AvgNa_mg.L)) +
  geom_point() +
  geom_line() +
  xlab("Date") +
  ylab("Average Sodium (mg/L)") +
  geom_smooth(method = lm)
```

\newpage

We also constructed and decomposed the time series for our flow dataset to determine whether discharge could be affecting our salinization results. The decomposition shows that discharge in the Neuse River basin has a seasonal component and a trend (Figure 12). The seasonal MannKendall test shows that discharge exhibits a monotonic trend over time, with flow significantly increasing year after year (p-value = 3.368e-07) (Figure 13, Table 2). This increasing trend in flow may be masking increases in salt content in the river, as increased flow will dilute an increased salt content to the same concentration, hiding an increase in salt deposition to the river.

```{r Discharge time series and decomp, include=FALSE}
#Creating Time Series for flow data
NeuseFlow_ts <- ts(NeuseFlow_processed[[2]], frequency =365)
NeuseFlow_decomp <- stl(NeuseFlow_ts, s.window = "periodic")
```

```{r Plot of Discharge Time Series Decomposition, include=TRUE, echo=FALSE, message=FALSE, fig.align='center', fig.width = 6, fig.height = 4, fig.cap="Time Series Decomposition of Discharge in the Neuse River."}
plot(NeuseFlow_decomp, main = "Discharge")
```

```{r Plot of Discharge over Time with Trend, include=TRUE, echo=FALSE, message=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap="Discharge in the Neuse River does exhibit a monotonic trend over time."}
ggplot(NeuseFlow_processed, aes(x = ActivityStartDate, y = discharge_cfs)) +
  geom_line() +
  xlab("Date") +
  ylab("Discharge (cfs)") +
  geom_smooth(method = lm)
```

\newpage

---

Table 2: P-values for Seasonal Mann-Kendall tests.

Measure           | P-value
------------------|------------------------
Conductivity      | 0.25276
Calcium           | 0.1038
Magnesium         | 0.29151
Sodium            | 0.091911
Discharge         | 3.368e-07

```{r Testing for Stationarity, include=FALSE, echo=FALSE, message=FALSE}
#Conductivity
Neuse_seasonality_trends_Conductivity <- Kendall::SeasonalMannKendall(NeuseCond_ts)
Neuse_seasonality_trends_Conductivity
summary(Neuse_seasonality_trends_Conductivity)
#conductivity_trend <- trend::smk.test(NeuseCond_ts)
#summary(conductivity_trend)

#Calcium
Neuse_seasonality_trends_Conductivity_Ca <- Kendall::SeasonalMannKendall(NeuseCond_Calcium_ts)
Neuse_seasonality_trends_Conductivity_Ca
summary(Neuse_seasonality_trends_Conductivity_Ca)

#Magnesium
Neuse_seasonality_trends_Conductivity_Mg <- Kendall::SeasonalMannKendall(NeuseCond_magnesium_ts)
Neuse_seasonality_trends_Conductivity_Mg
summary(Neuse_seasonality_trends_Conductivity_Mg)

#Sodium
Neuse_seasonality_trends_Conductivity_Na <- Kendall::SeasonalMannKendall(NeuseCond_sodium_ts)
Neuse_seasonality_trends_Conductivity_Na
summary(Neuse_seasonality_trends_Conductivity_Na)


#Testing for seasonality within Flow data
Neuse_seasonality_trends_Flow <- Kendall::SeasonalMannKendall(NeuseFlow_ts)
Neuse_seasonality_trends_Flow
summary(Neuse_seasonality_trends_Flow)
#discharge_trend <- trend::smk.test(NeuseFlow_ts)
#summary(discharge_trend)
```

After running the the Seasonal MannKendall test, we have found no significant change in the Neuse River over time for conductivity (p-value = 0.25276)(Table 2). However, we were able to reject the Null for flow data, and determine there has been a significant increase in discharge over time (p-value = 3.368e-07) (Table 2).


\newpage

### Is there seasonality in specific conductivity and flow of the Neuse River?

We plotted each mineral by specific conductance and colored by the day of the year (DOY) to visualize any seasonal patterns within a year (Figure 14). If the specific conductance or the different minerals exhibited seasonality, we would expect to see clusters of the same color within the plot. Although our time series analysis showed a seasonal component, we see that the DOY coloring shows no patterns, and thus, specific conductance and the minerals of Ca, Mg, and Na do not have any seasonal trend over the year. However, we do see that Na shows the strongest correlation with specific conductance out of all the minerals shown (stronger positive trend). 

```{r Create DOY column, include=FALSE, echo=FALSE, message=FALSE}
NeuseWQ_DOY <- NeuseWQ_processed_pivotLonger %>%
  filter(Year >= 2013 & Year <= 2022) %>%
  mutate(DOY = yday(ActivityStartDate)) 
```

```{r Plot by DOY, include=TRUE, echo=FALSE, message=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap="Salts by Specific Conductivity over the Year"}
saltsvspc_plot2 <- ggplot(NeuseWQ_DOY, aes(x = SpC_uS.cm, y = ResultMeasureValue, color = DOY, shape = Mineral)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c() +
  labs(x = expression("Specific Conductance ("*mu*"S/cm)"), y = "Mineral Concentration (mg/L)", color = "Day of Year")
saltsvspc_plot2

```

We also plotted boxplots of specific conductance and discharge for each month to visualize any seasonality over a year (Figure 15). We would expect to see high values of specific conductance in winter months when salts are applied to roads before/after snowfall, and low values specific conductance values in the summer months when road salts are not applied. We do not see any strong seasonal specific conductance trends in these plots. However, viewing the boxplots in conjunction with discharge, we see that there is generally an inverse relationship between the two. As discharge is high in winter months, specific conductance is low, likely because the minerals are diluted. 

```{r Boxplots of Specific Conductance and Discharge, echo=FALSE, fig.align='center', fig.height=6, fig.width=6, message=FALSE, warning=FALSE, include=TRUE, fig.cap="Comparison of specific conductance and discharge over the months."}
SpC_boxplot <- ggplot(NeuseWQ_processed, aes(x = as.factor(Month), y = SpC_uS.cm)) +
  geom_boxplot() +
  labs(x = "Month", y = expression("Specific Conductance ("*mu*"S/cm)"))
Flow_boxplot <- ggplot(NeuseFlow_processed, aes(x = as.factor(Month), y = discharge_cfs)) +
  geom_boxplot() +
  scale_y_log10() +
  labs(x = "Month", y = expression("Discharge (cfs)"))

plot_grid(SpC_boxplot, Flow_boxplot, ncol=1, align = "hv")
```

\newpage


## Question 2: Is calcium, magnesium, or sodium the driver of specific conductance?
While there is not much seasonality in the specific conductance found at this site in the Neuse River basin, the salt ions measured can still be examined to see if one or another appears to be driving the conductivity. It appears as though sodium is the ion most closely associated with the specific conductivity found at this site, as this ion both most closely aligns with the patterns seen in the data and also is the most abundant ion.

```{r Visualize 2013-2021 Conductivity, include=FALSE, echo=FALSE, message=FALSE, fig.align='left', fig.cap="Conductivity in the Neuse River from 2013 through 2021."}
#plot specific conductivity over same shorter time period
SpCshort_plot <- ggplot(NeuseWQ_short) +
  geom_line(aes(x = Date_monthrounded, y = AvgSpC_uS.cm)) +
  labs(x = "Date", y = expression("Specific Conductivity"))
SpCshort_plot
```

```{r Visualize 2013-2021 WQ data, include=FALSE}
#plot calcium over same shorter time period as specific conductivity
Ca.plot <- ggplot(NeuseWQ_short) +
  geom_line(aes(x = Date_monthrounded, y = AvgCa_mg.L)) +
  labs(x = "", y = expression("Calcium (mg/L)"))
Ca.plot

#plot Mg over same shorter time period
Mg.plot <- ggplot(NeuseWQ_short) +
  geom_line(aes(x = Date_monthrounded, y = AvgMg_mg.L)) +
  labs(x = "", y = expression("Magnesium (mg/L"))
Mg.plot

#plot sodium over same shorter time period
Na.plot <- ggplot(NeuseWQ_short) +
  geom_line(aes(x = Date_monthrounded, y = AvgNa_mg.L)) +
  labs(x = "Date", y = expression("Sodium (mg/L)"))
Na.plot

```

```{r Visualise all three components together, include=FALSE, echo=FALSE, message=FALSE, fig.align='left', fig.cap="Calcium, Magnesium, and Sodium in the Neuse River from 2013 through 2021"}
#now lets look at all three salt ions over the same time period
plot_grid(Ca.plot, Mg.plot, Na.plot,
          ncol = 1,
          align = "hv")
```

```{r Visualise salts and conductivity, include=TRUE, echo=FALSE, fig.align='left', fig.cap="Specific Conductivity and Salts in the Neuse River from 2013 through 2021"}
#pivot_longer
NeuseWQ_short_pivotLonger <- pivot_longer(NeuseWQ_short, AvgCa_mg.L:AvgSpC_uS.cm, names_to = "Mineral", values_to = "ResultMeasureValue")

#plot spec cond and the salts on the same plot to look for patterns
SpC.v.salts_plot <- ggplot(NeuseWQ_short_pivotLonger) +
  geom_line(aes(x = Date_monthrounded, y = ResultMeasureValue, color = Mineral)) +
  scale_y_log10() +
  scale_color_viridis_d() +
  labs(x = "Date", y = "Concentration", color = "Mineral")
SpC.v.salts_plot
```

```{r Plot by minerals by specific conductance, include=TRUE, echo=FALSE, message=FALSE, fig.align='center', fig.width=6, fig.height=4, fig.cap="Salts by Specific Conductivity."}
saltsvspc_plot3 <- ggplot(NeuseWQ_DOY, aes(x = SpC_uS.cm, y = ResultMeasureValue, color = Mineral, shape = Mineral)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_d() +
  labs(x = expression("Specific Conductance ("*mu*"S/cm)"), y = "Mineral Concentration (mg/L)", color = "Mineral")
saltsvspc_plot3

```

**[[[[[add corrplot?? talk about actual lm]]]]]**

```{r Correlation b/t Salts and Specific Conductivity, echo=FALSE}
ggplot(NeuseWQ_short) +
  geom_point(aes(x = AvgSpC_uS.cm, y = AvgNa_mg.L, color = "Na")) +
  geom_point(aes(x = AvgSpC_uS.cm, y = AvgCa_mg.L, color = "Ca")) +
  geom_point(aes(x = AvgSpC_uS.cm, y = AvgMg_mg.L, color = "Mg")) +
  labs(x = "Specific Conductivity", y = "Variable")

Na_lm <- lm(NeuseWQ_short$AvgNa_mg.L ~ NeuseWQ_short$AvgSpC_uS.cm)
summary(Na_lm)
Ca_lm <- lm(NeuseWQ_short$AvgCa_mg.L ~ NeuseWQ_short$AvgSpC_uS.cm)
summary(Ca_lm)
Mg_lm <- lm(NeuseWQ_short$AvgMg_mg.L ~ NeuseWQ_short$AvgSpC_uS.cm)
summary(Mg_lm)


```


\newpage

# Summary and Conclusions
The Neuse River at Kinston does not show strong seasonality in salinity and appears to show stationarity in salinity levels over the study period. Specific conductance, the proxy used to examine salinity, showed marginal seasonal trends when examined with a time series decomposition and no discernible pattern when plotted by the day of the year. Discharge similarly shows some marginal seasonality, but also shows a slight increasing trend. This slight increasing trend is interesting because it could hide a similarly-scaled increase in salinity over the same time period, as increased discharge would dilute more salinity effectively. 

Sodium appears to have the strongest impact on specific conductance, indicating that it perhaps is the main driver of the salinity that does exist in the Neuse River. Sodium is the most abundant ion in the water, magnesium and calcium had a much lower concentration. Sodium also appears to have a nearly 1:1 correlation with specific conductivity, as seen in the DOY plot where they are plotted against each other. 

Overall, this is good news for the Neuse, indicating that at Kinston, which is relatively far down the watershed, upstream uses of salts do not cause seasonal changes in the river. The location of Kinston in the watershed has been brought up in this analysis multiple times because it is significant. This gage site is potentially far enough downstream from where road salt would be used to have the salinity be insignificant by the time the water reached Kinston. This gage was chosen in part because it actually measures the salinity data we were interested in (and in part because we knew it was a pretty good dataset), but a future study would be interesting to compare the data at Kinston to similarly collected data at a gage site further upstream, like at Clayton, which is relatively close to the major metro area around Raleigh where road salt is also likely to be more prevalent than in smaller population areas. 
